---
title: "Assignment 3"
author: "Stefano Graziosi, Gabriele Molè, Laura Lo Schiavo, Giovanni Carron"
format: pdf
editor: visual
---

```{r echo = "FALSE"}
#| label: Load the relevant libraries

# Time series
library(dlm)
library(TSstudio)
library(feasts)
library(tseries)
  # Necessary packages for quantmod
  library(zoo)
  library(xts)
library(quantmod)

#Specifically for Assignment 2-3
library(depmixS4)
library(HiddenMarkov)

# Datasets
library(readr)
library(fpp3)

# For fancy plots
library(ggthemes)
  # Necessary packages for viridis
  library(viridisLite)
library(viridis)
library(gridExtra)
library(magrittr)
library(textab)

# Packages related to tidyverse, for data manipulation
library(tidyverse) # includes (lubridate), (dplyr), (ggplot2), (tidyr), (tidyselect)
library(tinytex)

# To handle time changes
library(timechange)


# To solve conflicts
library(conflicted)
conflicts_prefer(dplyr::filter)
```

```{r}
#| label: Load data
data <- read.csv("gistemp.txt", header = TRUE) 
monthly_data <- data[, 2:13] # Extract columns for Jan to Dec 
ts_data <- ts(as.vector(t(monthly_data)), start = c(1880, 1), frequency = 12)
```

# Research Question

> Consider the GISTEMP data, already used in Assignment 1. Can you envisage a latent signal (representing a latent trend) and read change-points in the temperature behavior along the years?

# Implementation

> Before proceeding with the analysis, obtain the seasonally adjusted time series (i.e. remove the seasonality).

```{r}

decompose_a_data <- decompose(ts_data, type = "additive")
decompose_m_data <- decompose(ts_data, type = "multiplicative")
stl_data <- stl(ts_data, s.window = "periodic")


# Remove seasonal component

seasonal_stl <- stl_data$time.series[, "seasonal"]


deseasoned_a <- ts_data - decompose_a_data$seasonal  
deseasoned_m <- ts_data - decompose_m_data$seasonal  
deseasoned_stl <- ts_data - seasonal_stl

# Subset to 1975
deseasoned_a_data_1975 <- window(deseasoned_a, end = c(1975, 12))
deseasoned_m_data_1975 <- window(deseasoned_m, end = c(1975, 12))
deseasoned_stl_1975 <- window(deseasoned_stl, end = c(1975, 12))

# Plot
#plot(deseasoned_a_data, main = "Deseasoned GISTEMP Data, additive decomposition")
plot(deseasoned_m, main = "Deseasoned GISTEMP Data, multiplicative decomposition")
plot(deseasoned_stl, main = "Deseasoned GISTEMP Data, stl decomposition")

#plot(deseasonalized_a_data_1975, main = "Deseasonalized GISTEMP Data (1880-1975), additive decomposition")
#plot(deseasonalized_m_data_1975, main = "Deseasonalized GISTEMP Data (1880-1975), multiplicative decomposition")
plot(deseasoned_stl_1975, main = "Deseasoned GISTEMP Data GISTEMP Data (1880-1975), stl decomposition")

```

# Hidden Markov Models

> Now explore the use of Hidden Markov Models to answer to the research question. Briefly support the choice of this model to answer the research question. Discuss (briefly but clearly) results and limitations.

> You might try an HHM with:

## State-dependent level of the temperature

> i.e. Y_t \| S_t = j \sym N(\mu\_j, \sigma\_j\^2) and decode the signal from the data Fit a 2-state HMM (e.g., "normal" vs. "warming" regimes)

```{r}

# Fit a 2-state HMM (e.g., "normal" vs. "warming" regimes)

y <- as.numeric(deseasoned_stl)
model <- depmix(y ~ 1, nstates = 2, data = data.frame(y), family = gaussian())
fmodel <- fit(model)

# Get the estimated state for each timestep
estStates <- posterior(fmodel)

# Plot
#plot(time(deseasoned_stl), estStates[,1], cex = .3)

i= estStates[1,1]
ii= if(i==1){i+1} else {i-1}
estMean1=fmodel@response[[i]][[1]]@parameters$coefficients
estMean2=fmodel@response[[ii]][[1]]@parameters$coefficients
estMeans=rep(estMean1, length(deseasoned_stl))
estMeans[estStates[,1]==ii]=estMean2
plot(deseasoned_stl)
title(main="Nile data and HMM estimated means", cex.main=1)
points(time(deseasoned_stl), estMeans, col="blue", cex=.3)

```

> Fit a 3-state HMM (e.g., "normal" vs. "warming" regimes)

```{r}

y <- as.numeric(deseasoned_stl)
model <- depmix(y ~ 1, nstates = 3, data = data.frame(y), family = gaussian())
fmodel <- fit(model)

# Get the estimated state for each timestep
estStates <- posterior(fmodel)

# Plot
plot(time(deseasoned_stl), estStates[,1], cex = .3)

n_states <- fmodel@nstates
estMeans <- numeric(length(deseasoned_stl))
for (s in 1:n_states) {
  state_mean <- fmodel@response[[s]][[1]]@parameters$coefficients
  estMeans[estStates[,1] == s] <- state_mean
}
plot(deseasoned_stl, main = "HMM Regime Means (Multiple States)")
points(time(deseasoned_stl), estMeans, col = "blue", cex = 0.3)

```

## up to 1975

> Fit a 2-state HMM (e.g., "normal" vs. "warming" regimes)

```{r}

# Fit a 2-state HMM (e.g., "normal" vs. "warming" regimes)

y <- as.numeric(deseasoned_stl_1975)
model <- depmix(y ~ 1, nstates = 2, data = data.frame(y), family = gaussian())
fmodel <- fit(model)

# Get the estimated state for each timestep
estStates <- posterior(fmodel)

# Plot
#plot(time(deseasoned_stl), estStates[,1], cex = .3)

i= estStates[1,1]
ii= if(i==1){i+1} else {i-1}
estMean1=fmodel@response[[i]][[1]]@parameters$coefficients
estMean2=fmodel@response[[ii]][[1]]@parameters$coefficients
estMeans=rep(estMean1, length(deseasoned_stl_1975))
estMeans[estStates[,1]==ii]=estMean2
plot(deseasoned_stl_1975)
title(main="Nile data and HMM estimated means", cex.main=1)
points(time(deseasoned_stl_1975), estMeans, col="blue", cex=.3)


```

> Fit a 3-state HMM (e.g., "normal" vs. "warming" regimes)

```{r}

y <- as.numeric(deseasoned_stl_1975)
model <- depmix(y ~ 1, nstates = 3, data = data.frame(y), family = gaussian())
fmodel <- fit(model)

# Get the estimated state for each timestep
estStates <- posterior(fmodel)

# Plot
#plot(time(deseasoned_stl_1975), estStates[,1], cex = .3)

n_states <- fmodel@nstates
estMeans <- numeric(length(deseasoned_stl_1975))

for (s in 1:n_states) {
  state_mean <- fmodel@response[[s]][[1]]@parameters$coefficients
  estMeans[estStates[,1] == s] <- state_mean
}
plot(deseasoned_stl_1975, main = "HMM Regime Means (Multiple States)")
points(time(deseasoned_stl_1975), estMeans, col = "blue", cex = 0.3)

```

## State-dependent linear trend

> i.e. Y_t \| S_t = j \sym N(\alpha\_j + \beta\_jt, \sigma\_j\^2)

```{r}

# Convert the deseasoned STL data to a data frame
deseasoned_df <- data.frame(
  temp_anomaly = as.numeric(deseasoned_stl),
  time = 1:length(deseasoned_stl)
)

# Remove NA values (if any at the ends from STL decomposition)
deseasoned_df <- na.omit(deseasoned_df)

# Fit HMM with 2 states and a state-dependent linear trend
n_states <- 3  # Try 3 as well to compare

set.seed(123)  # For reproducibility

mod <- depmix(temp_anomaly ~ time, 
              data = deseasoned_df, 
              nstates = n_states, 
              family = gaussian())

fit_mod <- fit(mod, verbose = TRUE)

# View model summary
summary(fit_mod)

# Posterior probabilities and state sequence
posterior_probs <- posterior(fit_mod)
deseasoned_df$state <- posterior_probs$state

ggplot(deseasoned_df, aes(x = time, y = temp_anomaly, color = factor(state))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Hidden State",
       x = "Time (months since Jan 1880)",
       y = "Deseasoned Temperature Anomaly") +
  theme_minimal()





```

> Fit a 3-state HMM (e.g., "normal" vs. "warming" regimes)

```{r}

# Convert the deseasoned STL data to a data frame
deseasoned_df <- data.frame(
  temp_anomaly = as.numeric(deseasoned_stl),
  time = 1:length(deseasoned_stl)
)

# Remove NA values (if any at the ends from STL decomposition)
deseasoned_df <- na.omit(deseasoned_df)

# Fit HMM with 2 states and a state-dependent linear trend
n_states <- 3  # Try 3 as well to compare

set.seed(123)  # For reproducibility

mod <- depmix(temp_anomaly ~ time, 
              data = deseasoned_df, 
              nstates = n_states, 
              family = gaussian())

fit_mod <- fit(mod, verbose = TRUE)

# View model summary
summary(fit_mod)

# Posterior probabilities and state sequence
posterior_probs <- posterior(fit_mod)
deseasoned_df$state <- posterior_probs$state

ggplot(deseasoned_df, aes(x = time, y = temp_anomaly, color = factor(state))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Hidden State",
       x = "Time (months since Jan 1880)",
       y = "Deseasoned Temperature Anomaly") +
  theme_minimal()

```

```{r}
# Convert the deseasoned STL data to a data frame
deseasoned_df <- data.frame(
  temp_anomaly = as.numeric(deseasoned_stl_1975),
  time = 1:length(deseasoned_stl_1975)
)

# Remove NA values (if any at the ends from STL decomposition)
deseasoned_df <- na.omit(deseasoned_df)

# Fit HMM with 2 states and a state-dependent linear trend
n_states <- 2  # Try 3 as well to compare

set.seed(123)  # For reproducibility

mod <- depmix(temp_anomaly ~ time, 
              data = deseasoned_df, 
              nstates = n_states, 
              family = gaussian())

fit_mod <- fit(mod, verbose = TRUE)

# View model summary
summary(fit_mod)

# Posterior probabilities and state sequence
posterior_probs <- posterior(fit_mod)
deseasoned_df$state <- posterior_probs$state

ggplot(deseasoned_df, aes(x = time, y = temp_anomaly, color = factor(state))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Hidden State",
       x = "Time (months since Jan 1880)",
       y = "Deseasoned Temperature Anomaly") +
  theme_minimal()
```

```{r}
# Convert the deseasoned STL data to a data frame
deseasoned_df <- data.frame(
  temp_anomaly = as.numeric(deseasoned_stl_1975),
  time = 1:length(deseasoned_stl_1975)
)

# Remove NA values (if any at the ends from STL decomposition)
deseasoned_df <- na.omit(deseasoned_df)

# Fit HMM with 2 states and a state-dependent linear trend
n_states <- 3  # Try 3 as well to compare

set.seed(123)  # For reproducibility

mod <- depmix(temp_anomaly ~ time, 
              data = deseasoned_df, 
              nstates = n_states, 
              family = gaussian())

fit_mod <- fit(mod, verbose = TRUE)



# Posterior probabilities and state sequence
posterior_probs <- posterior(fit_mod)
deseasoned_df$state <- posterior_probs$state


# Plot
ggplot(deseasoned_df, aes(x = time, y = temp_anomaly, color = factor(state))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(color = "Hidden State",
       x = "Time (months since Jan 1880)",
       y = "Deseasoned Temperature Anomaly") +
  theme_minimal()
```

## Case 3

```{r}
# Create lagged version of temp_anomaly
deseasoned_df <- deseasoned_df %>%
  mutate(temp_anomaly_lag1 = dplyr::lag(temp_anomaly)) %>%
  na.omit()  # Drop first NA row

# Set number of states
n_states <- 2

set.seed(123)

# Fit model with lag and time as covariates
mod_ar <- depmix(temp_anomaly ~ temp_anomaly_lag1 + time, 
                 data = deseasoned_df, 
                 nstates = n_states, 
                 family = gaussian())

fit_ar <- fit(mod_ar, verbose = TRUE)

# Add posterior state estimates
posterior_probs <- posterior(fit_ar)
deseasoned_df$state <- posterior_probs$state

# Convert the state variable to a factor (in case it's not already)
deseasoned_df$state <- as.factor(deseasoned_df$state)

# Create an empty plot with grey points
p <- ggplot(deseasoned_df, aes(x = time, y = temp_anomaly)) +
  geom_point(color = "grey60", alpha = 0.5) +
  labs(color = "Hidden State",
       x = "Time (months since Jan 1880)",
       y = "Deseasoned Temp Anomaly") +
  theme_minimal()

# Define colors for each state (adjust the number of colors based on the states)
state_colors <- scales::hue_pal()(length(unique(deseasoned_df$state)))

# Loop over each state and fit a linear regression for the entire time series
for (s in levels(deseasoned_df$state)) {
  state_data <- deseasoned_df %>%
    filter(state == s)
  
  # Fit the model for each state
  fit <- lm(temp_anomaly ~ temp_anomaly_lag1 + time, data = state_data)
  state_data$pred <- predict(fit, newdata = state_data)
  
  # Add the regression line for the current state to the plot
  p <- p + geom_line(data = state_data,
                     aes(x = time, y = pred),
                     color = state_colors[as.numeric(s)],  # Color by state
                     size = 0.8)
}

# Show the plot
print(p)
```

## Case 4

## Case 4
