---
title: "Assignment 3"
author: "Stefano Graziosi, Gabriele Molè, Laura Lo Schiavo, Giovanni Carron"
format: pdf
editor: visual
---

```{r echo = "FALSE"}
#| label: Load the relevant libraries

# Time series
library(dlm)
library(TSstudio)
library(feasts)
library(tseries)
  # Necessary packages for quantmod
  library(zoo)
  library(xts)
library(quantmod)

#Specifically for Assignment 2
library(depmixS4)
library(HiddenMarkov)

# Datasets
library(readr)
library(fpp3)

# For fancy plots
library(ggthemes)
  # Necessary packages for viridis
  library(viridisLite)
library(viridis)
library(gridExtra)
library(magrittr)
library(textab)

# Packages related to tidyverse, for data manipulation
library(tidyverse) # includes (lubridate), (dplyr), (ggplot2), (tidyr), (tidyselect)
library(tinytex)

# To handle time changes
library(timechange)


# To solve conflicts
library(conflicted)
conflicts_prefer(dplyr::filter)
```

```{r}
#| label: Load data
data <- read.csv("gistemp.txt", header = TRUE) 
monthly_data <- data[, 2:13] # Extract columns for Jan to Dec 
ts_data <- ts(as.vector(t(monthly_data)), start = c(1880, 1), frequency = 12)
```

# Research Question

> Consider the GISTEMP data, already used in Assignment 1. Can you envisage a latent signal (representing a latent trend) and read change-points in the temperature behavior along the years?

Ciao sono Laura

# Implementation

> Before proceeding with the analysis, obtain the seasonally adjusted time series (i.e. remove the seasonality).

# Hidden Markov Models

> Now explore the use of Hidden Markov Models to answer to the research question. Briefly support the choice of this model to answer the research question. Discuss (briefly but clearly) results and limitations.

## State-dependent level of the temperature

## State-dependent linear trend

## Case 3

## Case 4

## Case 4
