---
title: "Assignment 4 | Dynamic Linear Models with R"
author: "Stefano Graziosi"
format: html
editor: visual
---

```{r echo = "FALSE"}
#| label: Load the relevant libraries

# Time series
library(dlm)
library(TSstudio)
library(feasts)
library(tseries)
  # Necessary packages for quantmod
  library(zoo)
  library(xts)
library(quantmod)

#Specifically for Assignment 2
library(depmixS4)
library(HiddenMarkov)

# Datasets
library(readr)
library(fpp3)

# For fancy plots
library(ggthemes)
  # Necessary packages for viridis
  library(viridisLite)
library(viridis)
library(gridExtra)
library(magrittr)
library(textab)

# Packages related to tidyverse, for data manipulation
library(tidyverse) # includes (lubridate), (dplyr), (ggplot2), (tidyr), (tidyselect)
library(tinytex)

# To handle time changes
library(timechange)


# To solve conflicts
library(conflicted)
conflicts_prefer(dplyr::filter)
```

# Kalman lter for the random walk plus noise model

Consider the Nile data (measurements of the annual ow of the river Nile at Ashwan 1871-1970), available in R (`> ?Nile`).

First, plot the data. The series clearly appears non-stationary, presenting a quite evident change point. A local level model, i.e. a random walk plus noise, may be used to capture the main change point and other minor changes in the level of the Nile river. Let us consider the following random walk plus noise model.

$$
\begin{split}
y_t & = \theta_t + v_t \quad \quad v_t \sim \mathcal{N}(0,V) \\
\theta_t & = \theta_{t-1} + w_t \quad w_t \sim \mathcal{N}(0,W)
\end{split}
$$

with the due assumptions.

To start with, assume that the variances are known, $V = 15100$ , $W = 1470$. In fact, they will have to be estimated (next assignment). As the initial distribution, let $\theta_0 \sim \mathcal{N} (1000,1000)$.

```{r}
df <- Nile

nile_df <- data.frame(
  Y = as.numeric(time(Nile)),
  Flow = as.numeric(Nile)
)
```

```{r}
ggplot(nile_df, aes(x = Y, y = Flow)) +
  geom_line(color = "orange", size = 1) +     # Draw the line for the time series
  geom_point(color = "darkred", size = 1.2) +        # Add points to highlight each annual measurement
  labs(
    title = "Annual Flow of the Nile River",
    subtitle = "Data from the built-in R 'Nile' dataset",
    x = "Year",
    y = "Flow (10^3 cubic meters per second)"
  ) +
  theme_minimal() +                                # Use a minimal theme for a clean appearance
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    axis.text = element_text(color = "black")
  )
```

## Step 1: Filtering

> 1.  Compute the filtering estimates

```{r}
dlm_s1 <- dlm(m0 = 1000, C0 = 1000, FF = 1, V = 15100, GG = 1, W = 1470)

filtered_s1 <- dlmFilter(Nile, dlm_s1)
m_filtered_s1 <- dropFirst(filtered_s1$m)
```

> 2.  Compute the corresponding standard deviations and plot them. Comment briefly.

```{r}
Ct_list <- dlmSvd2var(filtered_s1$U.C, filtered_s1$D.C)
sd_filtered_s1 <- sqrt(unlist(Ct_list))[-1]

upper_95 <- m_filtered_s1 + 1.96 * sd_filtered_s1
lower_95 <- m_filtered_s1 - 1.96 * sd_filtered_s1
```

> 3.  Finally, plot the data together with the ltering state estimates and their 0.95 credible intervals.

```{r}
df_nile <- data.frame(
  Year      = as.numeric(time(Nile)),  
  Flow      = as.numeric(Nile),
  m_filt    = m_filtered_s1,
  sd_filt   = sd_filtered_s1,
  upper_95  = upper_95,
  lower_95  = lower_95
)

ggplot(df_nile, aes(x = Year)) +
  geom_line(aes(y = Flow), color = "orange", size = 1) + # Original data (gray line)
  geom_line(aes(y = m_filtered_s1), color = "steelblue", size = 1) + # Filtered estimates (blue line)
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), # 95% credibility interval around the filtered estimates
              fill = "steelblue", alpha = 0.2) +
  labs(
    title = "Local-Level Model Filtering on Nile Data",
    subtitle = "Kalman Filter Estimates with 95% Credible Intervals",
    x = "Year",
    y = "Annual Flow"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic")
  )
```

## Step 2: Online forecasting

> 1.  Compute the one-step ahead forecasts

```{r}
forecast_means <- dropFirst(filtered_s1$a)  # one-step ahead forecast means

forecast_var <- dlmSvd2var(filtered_s1$U.R, filtered_s1$D.R)
forecast_sd <- sqrt(unlist(forecast_var))[-1]
```

```{r}
upper_forecast <- forecast_means + 1.96 * forecast_sd
lower_forecast <- forecast_means - 1.96 * forecast_sd
```

> 2.  Plot the data, together the one-step-ahead forecasts and their 0.95 credible intervals

```{r}
df_forecast <- data.frame(
  Year      = as.numeric(time(Nile))[-1], 
  Flow      = as.numeric(Nile)[-1],
  Forecast  = forecast_means,
  sd_Forecast = forecast_sd,
  Lower_95 = lower_forecast,
  Upper_95 = upper_forecast
)

# Plot the one-step forecasts and the corresponding 95% credible intervals.
ggplot(df_forecast, aes(x = Year)) +
  geom_line(aes(y = Flow), color = "orange", size = 1) +
  geom_line(aes(y = Forecast), color = "darkgreen", size = 1) +
  geom_ribbon(aes(ymin = Lower_95, ymax = Upper_95), fill = "darkgreen", alpha = 0.2) +
  labs(
    title = "One-step Ahead Forecasts on Nile Data",
    subtitle = "Forecasts derived from the local-level (random walk plus noise) model",
    x = "Year",
    y = "Annual Flow"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic")
  )
```

## Step 3: Signal-to-noise Ratio

> What is the eect of the signal-to-noise ratio (i.e. the ratio $W/V$) on the forecasts? Repeat the exercise with dierent choices of V (observation variance) and W (evolution variance) and comment briey.

```{r}

scenarios <- data.frame(
  Scenario = c("Low ratio (W/V = 0.097)", "Unit ratio (W/V = 1)", "High ratio (W/V ≈ 10.3)"),
  V = c(15100, 15100, 1470),
  W = c(1470, 15100, 15100)
)

forecast_list <- list()

for(i in 1:nrow(scenarios)) {
  V_val <- scenarios$V[i]
  W_val <- scenarios$W[i]
  scenario_name <- scenarios$Scenario[i]
  
  dlm_model <- dlm(m0 = 1000, C0 = 1000, FF = 1, V = V_val, GG = 1, W = W_val)
  filter_mod <- dlmFilter(Nile, dlm_model)
  
  # Compute one-step forecasts
  f_mean <- dropFirst(filter_mod$a)
  f_var <- dlmSvd2var(filter_mod$U.R, filter_mod$D.R)
  f_sd <- forecast_sd <- sqrt(unlist(f_var))[-1]
  
  scenario_df <- data.frame(
    Year = as.numeric(time(Nile))[-1],
    Flow = as.numeric(Nile)[-1],
    Forecast = f_mean,
    sd_Forecast = f_sd,
    Lower_95 = f_mean - 1.96 * f_sd,
    Upper_95 = f_mean + 1.96 * f_sd,
    Scenario = scenario_name
  )
  
  forecast_list[[i]] <- scenario_df
}
```

```{r}
#| fig-width: 12
#| fig-height: 12
df_forecast_all <- bind_rows(forecast_list)

ggplot(df_forecast_all, aes(x = Year)) +
  geom_line(aes(y = Flow), color = "orange", size = 0.5) +
  geom_line(aes(y = Forecast), color = "darkgreen", size = 0.5) +
  geom_ribbon(aes(ymin = Lower_95, ymax = Upper_95), fill = "darkgreen", alpha = 0.2) +
  facet_wrap(~ Scenario, ncol = 1) +
  labs(
    title = "One-step Ahead Forecasts under Different Signal-to-Noise Ratios",
    subtitle = "Effect of varying W/V on forecasts",
    x = "Year",
    y = "Annual Flow"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic")
  )
```

## Step 4: Smoothing

> So far, for computations, we pretended that the data arrived sequentially. Now consider $(y_1, \ldots,y_T)$ and provide and plot the smoothing estimate of the Nile level $\theta_t$ at time t = 28 together with its 95% credible interval.

```{r}
smoothed <- dlmSmooth(Nile, dlm_s1)

theta_smoothed <- dropFirst(smoothed$s)

Ct_smooth_list <- dlmSvd2var(smoothed$U.S, smoothed$D.S)

var_smoothed <- unlist(Ct_smooth_list)[-1]
sd_smoothed <- sqrt(var_smoothed)

theta_t28 <- theta_smoothed[28]
sd_t28 <- sd_smoothed[28]
upper_t28 <- theta_t28 + 1.96 * sd_t28
lower_t28 <- theta_t28 - 1.96 * sd_t28

df_smooth <- data.frame(
  Year = as.numeric(time(Nile)),
  Flow = as.numeric(Nile),
  Smoothed = theta_smoothed,
  Lower_95 = theta_smoothed - 1.96 * sd_smoothed,
  Upper_95 = theta_smoothed + 1.96 * sd_smoothed
)
```

```{r}
#| fig-width: 12
year_t28 <- as.numeric(time(Nile))[28]

ggplot() +
  geom_line(data = df_smooth, aes(x = Year, y = Flow), color = "orange", size = 0.5) +
  geom_line(data = df_smooth, aes(x = Year, y = Smoothed), color = "steelblue", size = 0.5) +
  geom_ribbon(data = df_smooth, aes(x = Year, ymin = Lower_95, ymax = Upper_95), fill = "steelblue", alpha = 0.2) +
  # Highlight the smoothed estimate at t = 28 with error bars.
  geom_point(data = df_smooth %>% filter(Year == year_t28), aes(x = Year, y = Smoothed), color = "firebrick", size = 1.2) +
  geom_errorbar(data = df_smooth %>% filter(Year == year_t28), aes(x = Year, ymin = Lower_95, ymax = Upper_95), color = "firebrick", width = 0.5) +
  labs(
    title = "Smoothing Estimate for Nile River at t = 28",
    subtitle = paste0("Year ", year_t28, ": Smoothed level with 95% credible interval"),
    x = "Year",
    y = "Latent Level"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic")
  )
```
